---
title: "Combined analysis"
author: "David Tatarakis"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
Packages <-
  c(
    "dplyr",
    "Seurat",
    "aws.s3",
    "Matrix",
    "ggplot2",
    "pheatmap",
    "RColorBrewer",
    "SingleCellExperiment",
    "scuttle"
  )
lapply(Packages, library, character.only = TRUE)

`%notin%` <- Negate(`%in%`)

set.seed(1234)

diss_stress <-
  read.csv("/inst/extdata/diss_stress.csv",
           sep = ",",
           header = T)
```

```{r, include=FALSE}
microglia_Study1 <- readRDS("~/Study1_microglia_subset.Rds")

microglia_Study2 <-
  readRDS("~/ATV-Trem2_APPKI_microglia_analyzed.Rds")
```

So let's take a look at both datasets.

In study #1, we have now `r dim(microglia_Study1)[2]` microglia.

In study #2, we have `r dim(microglia_Study2)[2]` microglia.


```{r, echo=FALSE, fig.width=10, fig.height=4}
DimPlot(microglia_Study1, label = T) + 
  DimPlot(microglia_Study2, group.by = "seurat_clusters", label = T)
```

```{r, echo=FALSE, fig.width=12, fig.height=4}
DimPlot(
  microglia_Study1,
  group.by = "group",
  shuffle = T,
  raster = T,
  raster.dpi = c(1024, 1024)
) + DimPlot(
  microglia_Study2,
  group.by = "group",
  label = F,
  shuffle = T,
  raster = T,
  raster.dpi = c(1024, 1024)
) & NoAxes()
```

```{r}
table(microglia_Study2$group, microglia_Study2$sample)
```

```{r}
table(microglia_Study1$group, microglia_Study1$sample_id)
```

For the purposes of his combined analysis, we'll drop the untreated mouse from study #1, since it is a single sample with no replicates. 
```{r}
microglia_Study1 <- subset(microglia_Study1, subset = sample_id != "NH36G")
```

Now we'll combine the two studies. We will reduce the size and equalize the effects of each replicate. We'll take 1400 cells at random from each mouse sample.
```{r}
microglia_Study2$sample_id <- microglia_Study2$sample
microglia_Study2$barcode <- rownames(microglia_Study2@meta.data)
microglia_Study2$study <- "Study2"
microglia_Study1$study <- "Study1"



cells_to_keep1 <-
  microglia_Study1@meta.data %>% 
  group_by(sample_id) %>% 
  slice_sample(n = 1400)

cells_to_keep2 <-
  microglia_Study2@meta.data %>% 
  group_by(sample_id) %>% 
  slice_sample(n = 1400)

microglia_Study1 <-
  subset(microglia_Study1, cells = cells_to_keep1$barcode)
microglia_Study2 <-
  subset(microglia_Study2, cells = cells_to_keep2$barcode)

```

We will combine these studies using integration, to remove any study-specific effects.
```{r}
microglia_Study1 <- NormalizeData(microglia_Study1)
microglia_Study1 <- FindVariableFeatures(microglia_Study1)

microglia_Study2 <- NormalizeData(microglia_Study2)
microglia_Study2 <- FindVariableFeatures(microglia_Study2)
```

```{r, eval = FALSE}
anchors <-
  FindIntegrationAnchors(
    list(microglia_Study1, 
         microglia_Study2), 
    anchor.features = 2000)

combined <- IntegrateData(anchors)
```

```{r}
combined <- readRDS("~/atvTrem2_combined_objects/combined_object.Rds")
```

Now we have a final object consisting of `r dim(combined)[2]` microglia.

```{r, eval=FALSE}
DefaultAssay(combined) <- "RNA"

combined <- combined %>% 
  NormalizeData(.) %>%
  FindVariableFeatures(.) %>%
  ScaleData(.)
  
DefaultAssay(combined) <- "integrated"

combined <- combined %>%
  ScaleData(.)

PCA_features <- rownames(combined@assays$RNA@scale.data)[which(rownames(combined@assays$RNA@scale.data) %notin% diss_stress$symbol)]
  
combined <- combined %>%  
  RunPCA(., features = PCA_features) %>%
  RunUMAP(., dims = 1:9) %>%
  FindNeighbors(., dims = 1:9) %>%
  FindClusters(., resolution = 0.3)

combined$group <- paste0(combined$study, "_", combined$group)


combined$group <- plyr::mapvalues(combined$group, 
                                  from = unique(combined$group),
                                  to = c("Study1_ATV4D9_day28",
                                         "Study1_ATV4D9_day1",
                                         "Study1_ATVRSV",
                                         "Study1_ATV4D9_day7",
                                         "Study1_ATV4D9_day14",
                                         "Study2_WT_ATVDNP02",
                                         "Study2_WT_ATV4D9",
                                         "Study2_APPKI_ATVDNP02",
                                         "Study2_APPKI_ATV4D9",
                                         "Study2_APPKI_4D9"))

combined$group <- factor(combined$group, levels = c("Study1_ATVRSV",
                                                    "Study1_ATV4D9_day1",
                                                    "Study1_ATV4D9_day7",
                                                    "Study1_ATV4D9_day14",
                                                    "Study1_ATV4D9_day28",
                                                    "Study2_WT_ATVDNP02",
                                                    "Study2_WT_ATV4D9",
                                                    "Study2_APPKI_ATVDNP02",
                                                    "Study2_APPKI_4D9",
                                                    "Study2_APPKI_ATV4D9"))

saveRDS(combined, "~/atvTrem2_combined_objects/combined_object.Rds")


clustercols = brewer.pal("Dark2", n = 6)[c(1,3:6)]
combined$clusters <- plyr::mapvalues(combined$seurat_clusters,
                                     from = c(0,1,2,3,4,5),
                                     to = c("Homeostatic",
                                            "Homeostatic",
                                            "cluster2",
                                            "cluster3",
                                            "cluster4",
                                            "cluster5"))
```

```{r, fig.width=5, fig.height=3.25}
DimPlot(combined, cols = brewer.pal("Dark2", n = 6))
```

```{r, fig.width=5, fig.height=3.25}
DimPlot(combined, group.by = "clusters",cols = clustercols)
```

```{r, fig.width=10, fig.height=4}
DimPlot(
  combined,
  split.by = "study",
  group.by = "clusters",
  cols = clustercols,
  raster = T,
  raster.dpi = c(1024, 1024)
) & NoAxes()
```

```{r, fig.width=10, fig.height=3.5}
DimPlot(combined, split.by = "group", ncol = 5, group.by = "clusters",cols = clustercols) + NoAxes()
```

```{r, include=FALSE}
totals <-
  as.data.frame((table(combined$clusters, combined$sample_id))) %>%
  group_by(Var2) %>%
  mutate(percent = Freq / sum(Freq) * 100)

colnames(totals) <- c("cluster", "sample_id", "Count", "Percent")

meta_slim <-
  combined@meta.data %>% select(c(group, sample_id, study)) %>% distinct(.)

totals <- totals %>%
  left_join(meta_slim, by = "sample_id")

totals$cluster <-
  factor(totals$cluster, levels = rev(levels(totals$cluster)))
```

```{r, fig.width=12, fig.height=4}
ggplot(totals, aes(x = sample_id, y = Percent, fill = cluster)) +
  geom_bar(stat = "summary") +
  theme_bw() +
  scale_fill_manual(values = rev(brewer.pal("Dark2", n = 6)[c(1, 3:6)])) +
  coord_flip() +
  facet_wrap( ~ group, scales = "free_y", ncol = 5)
```

```{r}
totals$group <-
  factor(totals$group, levels = rev(levels(totals$group)))
```


```{r}
ggplot(totals, aes(x = group, y = Percent, fill = cluster)) +
  geom_bar(stat = "summary") +
  theme_bw() +
  scale_fill_manual(values = rev(clustercols)) +
  coord_flip() + scale_y_continuous(expand = expansion())
```

```{r, fig.width=6, fig.height=2}
ggplot(totals, aes(x = group, y = Percent, fill = cluster)) +
  geom_bar(stat = "summary") +
  theme_bw() +
  scale_fill_manual(values = rev(clustercols)) +
  coord_flip() +
  facet_wrap( ~ study, scales = "free_y", ncol = 1)
```


```{r, fig.width=10, fig.height=3.5}
DefaultAssay(combined) <- "RNA"
FeaturePlot(
  combined,
  c(
    "Tmem119",
    "Cst7",
    "Apoe",
    "Lpl",
    "Pcna",
    "Mki67",
    "Cxcl10",
    "Aldoa"
  ),
  ncol = 4,
  order = F
) & viridis::scale_color_viridis() & NoAxes()
```


```{r}
signatures <-
  read.csv(
    "/inst/extdata/state_signatures.csv",
    sep = ",",
    header = T
  )

chen_sigs <- signatures %>% filter(collection == "ChenColonna2021")
chen_sigs <- split(chen_sigs$symbol, chen_sigs$name)
names(chen_sigs) <- gsub("-", ".", names(chen_sigs))

topics <- signatures %>% filter(collection == "topics")
topics <- split(topics$symbol, topics$name)
```

```{r}
for (i in 1:length(chen_sigs)) {
  combined <- AddModuleScore(
    combined,
    features = list(chen_sigs[[i]]),
    name = paste0("ChenColonna_", names(chen_sigs)[i])
  )
}
```

```{r, fig.width=10, fig.height=1.5}
FeaturePlot(
  combined,
  paste0("ChenColonna_", names(chen_sigs), "1"),
  ncol = 5,
  order = F,
  min.cutoff = 0
) & viridis::scale_color_viridis() & NoAxes()
```


```{r}
sample_annotation <-
  combined@meta.data %>% select(c(group, sample_id)) %>% distinct(.)

sample_annotation <-
  sample_annotation %>% group_by(group) %>% dplyr::arrange(.by_group = T) %>% as.data.frame()
rownames(sample_annotation) <- sample_annotation$sample_id

sample_annotation_slim <- sample_annotation %>% dplyr::select(group)
rownames(sample_annotation_slim) <- sample_annotation$sample_id
```

```{r}
chen_annotation <-
  signatures %>% dplyr::filter(collection == "ChenColonna2021")
rownames(chen_annotation) <- chen_annotation$symbol
chen_annotation <- dplyr::select(chen_annotation, name)

chen_annotation_colors <- list(
  name = c(
    `Homeostatic` = "lightblue3",
    `DAM-like` = "coral2",
    `IFN` = "darkgreen",
    `MHC` = "grey20",
    `Cyc-M` = "lemonchiffon3"
  ),
  group = c(
    `Study1_ATVRSV` = "#8DD3C7",
    `Study1_ATV4D9_day1` = "#FFFFB3",
    `Study1_ATV4D9_day7` = "#BEBADA",
    `Study1_ATV4D9_day14` = "#FB8072",
    `Study1_ATV4D9_day28` = "#80B1D3",
    `Study2_WT_ATVDNP02` = "#FDB462",
    `Study2_WT_ATV4D9` = "#B3DE69",
    `Study2_APPKI_ATVDNP02` = "#FCCDE5",
    `Study2_APPKI_4D9` = "#D9D9D9",
    `Study2_APPKI_ATV4D9` = "#BC80BD"
  )
)
```

```{r}
chen_averages <-
  AverageExpression(combined,
                    features = rownames(chen_annotation),
                    group.by = "sample_id")

chen_cluster_averages <-
  AverageExpression(combined,
                    features = rownames(chen_annotation),
                    group.by = "seurat_clusters")
```

```{r, echo=FALSE, fig.width=6.5, fig.height=8}
pheatmap::pheatmap(
  mat = chen_averages$RNA[, rownames(sample_annotation_slim)],
  scale = "row",
  annotation_row = chen_annotation,
  annotation_col = data.frame(sample_annotation_slim),
  annotation_colors = chen_annotation_colors,
  cluster_cols = F,
  cluster_rows = F,
  show_rownames = T,
  show_colnames = F,
  gaps_col = c(3, 6, 9, 12, 15, 19, 23, 27, 31),
  gaps_row = c(15, 40, 49, 55),
  main = "Chen/Colonna Review Signature Genes",
  color = rev(colorRampPalette(RColorBrewer::brewer.pal(11, "RdBu"))(256))
)
```


```{r, echo=FALSE, fig.width=4, fig.height=8}
pheatmap::pheatmap(
  mat = chen_cluster_averages$RNA,
  scale = "row",
  annotation_row = chen_annotation,
  annotation_colors = chen_annotation_colors[1],
  cluster_cols = F,
  cluster_rows = F,
  show_rownames = T,
  show_colnames = T,
  gaps_row = c(15, 40, 49, 55),
  main = "Chen/Colonna Review Signature Genes",
  color = rev(colorRampPalette(RColorBrewer::brewer.pal(11, "RdBu"))(256))
)
```
