---
title: "Microglia analysis"
author: "David Tatarakis"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
libraries <- c("aws.s3", "Seurat", "Matrix", "ggplot2", "dplyr", "RColorBrewer")
lapply(libraries, library, character.only=T)

`%notin%` <- Negate(`%in%`)
set.seed(1234)
```

# Dimensionality reduction
```{r}
# read in the integrated object from the prior vignette
microglia <- readRDS("~/atvTrem2_combined_objects/NGS000072_microglia.Rds")
```



```{r}
DefaultAssay(microglia) <- "RNA"

diss_stress <- read.csv("~/atvTrem2_combined_scRNAseq/inst/extdata/diss_stress.csv")

microglia <- microglia %>%
  NormalizeData(.) %>%
  FindVariableFeatures(.) %>%
  ScaleData(.)

# Run PCA on variable features without dissociation stress signature genes
PCA_features <- rownames(microglia@assays$RNA@scale.data)[which(rownames(microglia@assays$RNA@scale.data) %notin% diss_stress$symbol)]

microglia <- microglia %>% RunPCA(., features = PCA_features)

microglia <- microglia %>%
  RunUMAP(., dims = 1:8) %>%
  FindNeighbors(., dims = 1:8) %>%
  FindClusters(., resolution = 0.2)
```

```{r, fig.width=4.5, fig.height=3}
DimPlot(microglia)
```

```{r,fig.width=8.75, fig.height=2.5}
DimPlot(microglia, split.by = "batch")
```


We can see there is one small cluster (4) that represent low quality cells that got past our filtering. So now I'll remove those cells and repeat processing.
```{r}
VlnPlot(microglia, c("nCount_RNA", "nFeature_RNA"))
```

```{r}
microglia <- subset(microglia, subset = seurat_clusters != 4)
```


We now have `r dim(microglia)[2]` microglia. That's probably far more than we need, and it wil lmake the analysis very computationally intensive. So Let's trim the object down. We'll take 6000 cells from each condition for a total of 30,000 cells.

```{r}
microglia[["group"]] <- paste0(microglia$treatment, "_", "day", microglia$treatment_duration)

microglia$group <- plyr::mapvalues(microglia$group,
                                   from = c("No treatment_dayN/A", "ATV:RSV_day1"),
                                   to = c("Untreated", "ATV:RSV"))

microglia$group <- factor(microglia$group, levels = c("Untreated",
                                                      "ATV:RSV",
                                                      "ATV:4D9_day1",
                                                      "ATV:4D9_day7",
                                                      "ATV:4D9_day14",
                                                      "ATV:4D9_day28"))
```

```{r}
microglia[["barcode"]] <- rownames(microglia@meta.data)
cells_to_keep <- microglia@meta.data %>% group_by(group) %>% slice_sample(., n= 5000)
cells_to_keep <- cells_to_keep$barcode

microglia_subset <- subset(microglia, cells =cells_to_keep)
```

```{r}
microglia_subset <- microglia_subset %>%
  NormalizeData(.) %>%
  FindVariableFeatures(.) %>%
  ScaleData(.)

# Run PCA on variable features without dissociation stress signature genes
PCA_features <- rownames(microglia_subset@assays$RNA@scale.data)[which(rownames(microglia_subset@assays$RNA@scale.data) %notin% diss_stress$symbol)]

microglia_subset <- microglia_subset %>% RunPCA(., features = PCA_features)

microglia_subset <- microglia_subset %>%
  RunUMAP(., dims = 1:8) %>%
  FindNeighbors(., dims = 1:8) %>%
  FindClusters(., resolution = 0.6)
```

```{r, fig.width=4.5, fig.height=3}
DimPlot(microglia_subset, cols = c(rev(brewer.pal("Dark2", n = 8)), "dodgerblue3", "firebrick2", "gray90", "black"))
```

```{r,fig.width=8.75, fig.height=2.5}
DimPlot(microglia_subset, split.by = "batch", cols = c(rev(brewer.pal("Dark2", n = 8)), "dodgerblue3", "firebrick2", "gray90", "black"))
```

```{r,fig.width=8.75, fig.height=4.5}
DimPlot(microglia_subset, split.by = "group", ncol = 3, cols = c(rev(brewer.pal("Dark2", n = 8)), "dodgerblue3", "firebrick2", "gray90", "black"))
```

What genes define each of these clusters?
```{r, eval = FALSE}
markers <- FindAllMarkers(microglia_subset, only.pos = T, min.pct = 0.25)
write.csv(markers, file = "~/atvTrem2_combined_scRNAseq/inst/extdata/NGS000072_microglia_cluster_markers.csv")
```

```{r}
markers <- read.csv("~/atvTrem2_combined_scRNAseq/inst/extdata/NGS000072_microglia_cluster_markers.csv", header = T, row.names = 1)

top5 <- markers %>% group_by(cluster) %>% slice_max(n = 5, order_by = avg_log2FC)
```

```{r, fig.width=8, fig.height=3}
FeaturePlot(microglia_subset, c("Tmem119", "Cst7", "Pcna", "Mki67", "Cxcl10", "Ifitm3", "Isg15", "Apoe"), ncol = 4, order = T) & viridis::scale_color_viridis() & NoAxes()
```

```{r, eval = FALSE}
microglia_subset <- RenameIdents(microglia_subset, 
                                 `0` = ,
                                 `1` = ,
                                 `1` = ,
                                 `3` = ,
                                 `4` = ,
                                 `5` = ,
                                 `6` = ,
                                 `7` = ,
                                 `8` = ,
                                 `9` = ,
                                 `10` = ,
                                 `11` = ,) 
```

#cols = c(brewer.pal("Set1", n = 9), "gray40", "gray60", "gray70")
```{r, eval=FALSE}
saveRDS(microglia_subset, "~/atvTrem2_combined_objects/NGS000072_microglia_subset.Rds")
```
