---
title: "mouse_hTREM2_KI_bulk_analysis"
author: "David Tatarakis"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,message=FALSE, warning=FALSE}
library(FacileBiocData)
library(FacileAnalysis)
library(FacileIncubator)
library(edgeR)
library(dplyr)
library(tidyverse)
library(tidyr)
library(pheatmap)
library(ggplot2)
library(here)
```

```{r, message=FALSE, warning=FALSE}
# Read in the data
DGEList <- readRDS(here("inst/exdata/mouse_hTREM2_KI_bulk-rna-seq.rds"))
gene.all <- DGEList$genes
prot.genes <- gene.all %>% filter(meta == "protein_coding")

# Hallmarks gene signature lists 
gdb.h <- sparrow::getMSigGeneSetDb("H", "mouse")
```


```{r, message=FALSE, warning=FALSE}
DGEList_10mgpkg <- DGEList[,which(DGEList$samples$treatment_dose %in% 
                                    c("DNL919__mgkg10", 
                                      "Vehicle__mgkg00", 
                                      "ATV_DNP02__mgkg10"))]

DGEList_10mgpkg <- calcNormFactors(DGEList_10mgpkg)

# Filter lowly expressed genes and non protein-coding genes
keep <- edgeR::filterByExpr(y = DGEList_10mgpkg)
DGEList_filt <- DGEList_10mgpkg[which(keep), ]
DGEList_filt <- DGEList_filt[which(rownames(DGEList_filt) %in% 
                                     prot.genes$feature_id), ]

DGEList_filt$samples$group <- factor(DGEList_filt$samples$group, 
                                     levels = c("Vehicle",
                                                "ATV_DNP02",
                                                "DNL919__mgkg10__days_01",
                                                "DNL919__mgkg10__days_04",
                                                "DNL919__mgkg10__days_07"))

meta.data <- DGEList_filt$samples
```

```{r, message=FALSE, warning=FALSE}
group <- DGEList_filt$samples$group
batch <- DGEList_filt$samples$batch
mm <- model.matrix(~0 + group + batch)

y <- voom(DGEList_filt, mm, plot = T)
```

```{r, message=FALSE, warning=FALSE}
plotMDS(y$E, col = as.numeric(DGEList_filt$samples$group))
```





```{r, message=FALSE, warning=FALSE}
fit <- lmFit(y, mm)
contr <- makeContrasts(
  DNP02vsVeh = groupATV_DNP02 - groupVehicle,
  Day1vsVeh = groupDNL919__mgkg10__days_01 - groupVehicle,
  Day4vsVeh = groupDNL919__mgkg10__days_04 - groupVehicle,
  Day7vsVeh = groupDNL919__mgkg10__days_07 - groupVehicle,
  Day1vsDNP02 = groupDNL919__mgkg10__days_01 - groupATV_DNP02,
  Day4vsDNP02 = groupDNL919__mgkg10__days_04 - groupATV_DNP02,
  Day7vsDNP02 = groupDNL919__mgkg10__days_07 - groupATV_DNP02,                       
  levels = colnames(coef(fit)))


tmp <- contrasts.fit(fit, contr)
tmp <- eBayes(tmp)

### Perform pairwise comparisons for day 1, 4, and 7 against vehicle
DNP02vsVeh.top.table <- topTable(tmp, 
                                 sort.by = "P", 
                                 n = Inf, 
                                 coef = "DNP02vsVeh")
Day1vsVeh.top.table <- topTable(tmp, 
                                sort.by = "P", 
                                n = Inf, 
                                coef = "Day1vsVeh")
Day4vsVeh.top.table <- topTable(tmp, 
                                sort.by = "P", 
                                n = Inf, 
                                coef = "Day4vsVeh")
Day7vsVeh.top.table <- topTable(tmp, 
                                sort.by = "P", 
                                n = Inf, 
                                coef = "Day7vsVeh")

### Perform pairwise comparisons for day 1, 4, and 7 against DNP02
Day1vsDNP02.top.table <- topTable(tmp, 
                                  sort.by = "P", 
                                  n = Inf, 
                                  coef = "Day1vsDNP02")
Day4vsDNP02.top.table <- topTable(tmp, 
                                  sort.by = "P", 
                                  n = Inf, 
                                  coef = "Day4vsDNP02")
Day7vsDNP02.top.table <- topTable(tmp, 
                                  sort.by = "P", 
                                  n = Inf, 
                                  coef = "Day7vsDNP02")

```

# DE 10mg/kg vs Vehicle {.tabset}

```{r, message=FALSE, warning=FALSE}
# DNP02 vs Vehicle
DNP02vsVeh.top.table$diffex <- "NO"
DNP02vsVeh.top.table$diffex[DNP02vsVeh.top.table$logFC > 0.5 & 
                              DNP02vsVeh.top.table$adj.P.Val < 0.05] <- "UP"
DNP02vsVeh.top.table$diffex[DNP02vsVeh.top.table$logFC < -0.5 & 
                              DNP02vsVeh.top.table$adj.P.Val < 0.05] <- "DOWN"

p1 <- ggplot(data = DNP02vsVeh.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
      geom_point() + 
      theme_bw() + 
      xlim(c(-9, 9)) +
      ylim(c(0, 15)) +
      scale_color_manual(values = c("black")) + 
      geom_vline(xintercept = c(-0.5, 0.5), 
                 col = "gray40", 
                 linetype = "dashed") + 
      geom_hline(yintercept = -log10(0.05), 
                 col = "gray40", 
                 linetype = "dashed") + 
      theme(legend.position = "none") + 
      ggtitle("DNP02") +
      labs(x = "log2FC vs Vehicle", y = "-log10(adj p value)")


# Day1 vs Vehicle
Day1vsVeh.top.table$diffex <- "NO"
Day1vsVeh.top.table$diffex[Day1vsVeh.top.table$logFC > 0.5 & 
                             Day1vsVeh.top.table$adj.P.Val < 0.05] <- "UP"
Day1vsVeh.top.table$diffex[Day1vsVeh.top.table$logFC < -0.5 & 
                             Day1vsVeh.top.table$adj.P.Val < 0.05] <- "DOWN"

p2 <- ggplot(data = Day1vsVeh.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
      geom_point() + 
      theme_bw() + 
      xlim(c(-9, 9)) +
      ylim(c(0, 15)) +
      scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
      geom_vline(xintercept = c(-0.5, 0.5), 
                 col = "gray40", 
                 linetype = "dashed") + 
      geom_hline(yintercept = -log10(0.05), 
                 col = "gray40", 
                 linetype = "dashed") + 
      theme(legend.position = "none") + 
      ggtitle("Day 1 ATV:TREM2") +
      labs(x = "log2FC vs Vehicle", y = "-log10(adj p value)") 


# Day 4 vs Vehicle
Day4vsVeh.top.table$diffex <- "NO"
Day4vsVeh.top.table$diffex[Day4vsVeh.top.table$logFC > 0.5 & 
                             Day4vsVeh.top.table$adj.P.Val < 0.05] <- "UP"
Day4vsVeh.top.table$diffex[Day4vsVeh.top.table$logFC < -0.5 & 
                             Day4vsVeh.top.table$adj.P.Val < 0.05] <- "DOWN"


p3 <- ggplot(data = Day4vsVeh.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
      geom_point() + 
      theme_bw() + 
      xlim(c(-9, 9)) +
      ylim(c(0, 15)) +
      scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
      geom_vline(xintercept = c(-0.5, 0.5), 
                 col = "gray40", 
                 linetype = "dashed") + 
      geom_hline(yintercept = -log10(0.05), 
                 col = "gray40", 
                 linetype = "dashed") + 
      theme(legend.position = "none") + 
      ggtitle("Day 4 ATV:TREM2") +
      labs(x = "log2FC vs Vehicle", y = "-log10(adj p value)") 


# Day 7 vs Vehicle
Day7vsVeh.top.table$diffex <- "NO"
Day7vsVeh.top.table$diffex[Day7vsVeh.top.table$logFC > 0.5 & 
                             Day7vsVeh.top.table$adj.P.Val < 0.05] <- "UP"
Day7vsVeh.top.table$diffex[Day7vsVeh.top.table$logFC < -0.5 & 
                             Day7vsVeh.top.table$adj.P.Val < 0.05] <- "DOWN"


p4 <- ggplot(data = Day7vsVeh.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
      geom_point() + 
      theme_bw() + 
      xlim(c(-9, 9)) +
      ylim(c(0, 15)) +
      scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
      geom_vline(xintercept = c(-0.5, 0.5), 
                 col = "gray40", 
                 linetype = "dashed") + 
      geom_hline(yintercept = -log10(0.05), 
                 col = "gray40", 
                 linetype = "dashed") + 
      theme(legend.position = "none") + 
      ggtitle("Day 7 ATV:TREM2") +
      labs(x = "log2FC vs Vehicle", y = "-log10(adj p value)") 
```


```{r, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
```

## Vehicle
```{r, warning = FALSE, message= FALSE}
DT::datatable(DNP02vsVeh.top.table %>% 
                select(c(
                  name, 
                  logFC, 
                  t, 
                  P.Value, 
                  adj.P.Val)), 
              filter = "bottom")
```


## Day 1
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day1vsVeh.top.table %>% 
                select(c(
                  name, 
                  logFC, 
                  t, 
                  P.Value, 
                  adj.P.Val)), 
              filter = "bottom")
```

## Day 4
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day4vsVeh.top.table %>% 
                select(c(
                  name, 
                  logFC, 
                  t, 
                  P.Value, 
                  adj.P.Val)), 
              filter = "bottom")
```


## Day 7
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day7vsVeh.top.table %>% 
                select(c(
                  name, 
                  logFC, 
                  t, 
                  P.Value, 
                  adj.P.Val)), 
              filter = "bottom")
```


# DE 10mg/kg vs ATV:ISO {.tabset}

```{r, message=FALSE, warning=FALSE}
# Day 1 vs DNP02
Day1vsDNP02.top.table$diffex <- "NO"
Day1vsDNP02.top.table$diffex[Day1vsDNP02.top.table$logFC > 0.5 & 
                               Day1vsDNP02.top.table$adj.P.Val < 0.05] <- "UP"
Day1vsDNP02.top.table$diffex[Day1vsDNP02.top.table$logFC < -0.5 & 
                               Day1vsDNP02.top.table$adj.P.Val < 0.05] <- "DOWN"


p2 <- ggplot(data = Day1vsDNP02.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
          geom_point() + 
          theme_bw() + 
          xlim(c(-9, 9)) +
          ylim(c(0, 15)) +
          scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
          geom_vline(xintercept = c(-0.5, 0.5), 
                     col = "gray40", 
                     linetype = "dashed") + 
          geom_hline(yintercept = -log10(0.05), 
                     col = "gray40", 
                     linetype = "dashed") + 
          theme(legend.position = "none") + 
          ggtitle("Day 1 ATV:TREM2") +
          labs(x = "log2FC vs ATV:ISO", y = "-log10(adj p value)") 


# Day 4 vs DNP02
Day4vsDNP02.top.table$diffex <- "NO"
Day4vsDNP02.top.table$diffex[Day4vsDNP02.top.table$logFC > 0.5 & 
                               Day4vsDNP02.top.table$adj.P.Val < 0.05] <- "UP"
Day4vsDNP02.top.table$diffex[Day4vsDNP02.top.table$logFC < -0.5 & 
                               Day4vsDNP02.top.table$adj.P.Val < 0.05] <- "DOWN"


p3 <- ggplot(data = Day4vsDNP02.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
          geom_point() + 
          theme_bw() + 
          xlim(c(-9, 9)) +
          ylim(c(0, 15)) +
          scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
          geom_vline(xintercept = c(-0.5, 0.5), 
                     col = "gray40", 
                     linetype = "dashed") + 
          geom_hline(yintercept = -log10(0.05),
                     col = "gray40", 
                     linetype = "dashed") + 
          theme(legend.position = "none") + 
          ggtitle("Day 4 ATV:TREM2") +
          labs(x = "log2FC vs ATV:ISO", y = "-log10(adj p value)") 


# Day 7 vs DNP02
Day7vsDNP02.top.table$diffex <- "NO"
Day7vsDNP02.top.table$diffex[Day7vsDNP02.top.table$logFC > 0.5 & 
                               Day7vsDNP02.top.table$adj.P.Val < 0.05] <- "UP"
Day7vsDNP02.top.table$diffex[Day7vsDNP02.top.table$logFC < -0.5 & 
                               Day7vsDNP02.top.table$adj.P.Val < 0.05] <- "DOWN"


p4 <- ggplot(data = Day7vsDNP02.top.table, 
             aes(x=logFC, 
                 y = -log10(adj.P.Val), 
                 col = diffex, 
                 label = name)) +
          geom_point() + 
          theme_bw() + 
          xlim(c(-9, 9)) +
          ylim(c(0, 15)) +
          scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) + 
          geom_vline(xintercept = c(-0.5, 0.5), 
                     col = "gray40", 
                     linetype = "dashed") + 
          geom_hline(yintercept = -log10(0.05), 
                     col = "gray40", 
                     linetype = "dashed") + 
          theme(legend.position = "none") + 
          ggtitle("Day 7 ATV:TREM2") +
          labs(x = "log2FC vs ATV:ISO", y = "-log10(adj p value)") 
```


```{r, fig.width=7.5, fig.height=2.5, message=FALSE, warning=FALSE}
gridExtra::grid.arrange(p2, p3, p4, nrow = 1)
```


## Day 1
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day1vsDNP02.top.table %>% select(c(name, 
                                                 logFC, 
                                                 t, 
                                                 P.Value, 
                                                 adj.P.Val)), 
              filter = "bottom")
```

## Day 4
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day4vsVeh.top.table %>% select(c(name, 
                                               logFC, 
                                               t, 
                                               P.Value, 
                                               adj.P.Val)), 
              filter = "bottom")
```

## Day 7
```{r, warning = FALSE, message= FALSE}
DT::datatable(Day7vsVeh.top.table %>% select(c(name, 
                                               logFC, 
                                               t, 
                                               P.Value, 
                                               adj.P.Val)), 
              filter = "bottom")
```

# GSEA 10mg/kg vs Vehicle
```{r, message=FALSE, warning=FALSE}
day1gsea <- ffsea(x = Day1vsVeh.top.table, 
                  rank_by = "t",
                  methods = "fgsea", 
                  fsets = gdb.h)

day1gs_stats_all <- day1gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day1gs_stats <- day1gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    filter(padj < 0.05) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day1gs_stats <- day1gs_stats %>% 
  mutate(direction = ifelse(NES>0, "up", "down"))
```

```{r, message=FALSE, warning=FALSE}
DT::datatable(day1gs_stats_all %>% select(c(name, n, NES, padj)), 
              filter = "bottom", 
              caption = "Hallmark pathways affected at Day1 DNL 919 treatment")
```

```{r, fig.width=10,fig.height=4}
ggplot(day1gs_stats, aes(x = reorder(name, NES), y = NES, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_bw() +
  scale_fill_manual(values = c("dodgerblue3", "firebrick3")) +
  theme(legend.position = "none") +
  labs(x = "", y = "GSEA Normalized Enrichment Score")
  
```

# Metabolic pathways vs Vehicle
```{r, message=FALSE, warning=FALSE}
##Get log2(cpm) and sample info. Arrange samples by group.
norm.counts <- log2(cpm(DGEList_filt) + 1)
sample_annotation <- DGEList_filt$samples %>% arrange(group)


##Get the genes from each pathway in the GSEA
oxphos_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_OXIDATIVE_PHOSPHORYLATION"),]$leadingEdge %>% 
  unlist() 
glycolysis_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_GLYCOLYSIS"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
metabolism.counts <- norm.counts[c(oxphos_gsea_genes[2:16], 
                                   glycolysis_gsea_genes[1:15]), 
                                 rownames(sample_annotation)]
gene_names <- DGEList_filt$genes %>% select(c(feature_id, name))

##Change rownames to common gene symbols
rownames(metabolism.counts) <- gene_names[rownames(metabolism.counts),2]


gene_annotation <- data.frame(row.names = rownames(metabolism.counts), 
                              pathway = c(rep("OXPhos", 15), 
                                          rep("Glycolysis", 15)))
column_annotation <- data.frame(row.names = rownames(sample_annotation), 
                                group = sample_annotation$group)

annoCol<-list(pathway=c(OXPhos="Gray20", Glycolysis="Gray60"),
              group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = metabolism.counts,
                   cluster_rows = F,
                   cluster_cols = F,
                   scale= "row",
                   gaps_row = c(15),
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   annotation_row = gene_annotation,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "Metabolism DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```

# Other activation signatures vs Vehicle
```{r, message=FALSE, warning=FALSE}
##Get the genes from each pathway in the GSEA
allograft_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_ALLOGRAFT_REJECTION"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
allograft.counts <- norm.counts[allograft_gsea_genes[1:25], 
                                rownames(sample_annotation)]

##Change rownames to common gene symbols
rownames(allograft.counts) <- gene_names[rownames(allograft.counts),2]

annoCol2<-list(group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = allograft.counts,
                   cluster_rows = T,
                   cluster_cols = F,
                   scale= "row",
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   treeheight_row = 0,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "Allograft Rejection DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```


```{r, message=FALSE, warning=FALSE}
##Get the genes from each pathway in the GSEA
G2M_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_G2M_CHECKPOINT"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
G2M.counts <- norm.counts[G2M_gsea_genes[1:25], rownames(sample_annotation)]

##Change rownames to common gene symbols
rownames(G2M.counts) <- gene_names[rownames(G2M.counts),2]

annoCol2<-list(group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = G2M.counts,
                   cluster_rows = T,
                   cluster_cols = F,
                   scale= "row",
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   treeheight_row = 0,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "G2M Checkpoint DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```


# GSEA 10mg/kg vs ATV:ISO {.tabset}

## day 1
```{r, message=FALSE, warning=FALSE}
day1gsea <- ffsea(x = Day1vsVeh.top.table, 
                  rank_by = "t", 
                  methods = "fgsea", 
                  fsets = gdb.h)

day1gs_stats_all <- day1gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day1gs_stats <- day1gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    filter(padj < 0.05) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day1gs_stats <- day1gs_stats %>% 
  mutate(direction = ifelse(NES>0, "up", "down"))
```

```{r, message=FALSE, warning=FALSE}
DT::datatable(day1gs_stats_all %>% 
              select(c(name, n, NES, padj)), 
              filter = "bottom", 
              caption = "Hallmark pathways affected at 
              Day1 DNL 919 treatment vs ATV:ISO")
```

```{r, fig.width=10,fig.height=4}
ggplot(day1gs_stats, aes(x = reorder(name, NES), y = NES, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_bw() +
  scale_fill_manual(values = c("dodgerblue3", "firebrick3")) +
  theme(legend.position = "none") +
  labs(x = "", y = "GSEA Normalized Enrichment Score")
  
```

## day 4
```{r, message=FALSE, warning=FALSE}
day4gsea <- ffsea(x = Day4vsVeh.top.table, 
                  rank_by = "t", 
                  methods = "fgsea", 
                  fsets = gdb.h)

day4gs_stats_all <- day4gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day4gs_stats <- day4gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    filter(padj < 0.05) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day4gs_stats <- day4gs_stats %>% 
  mutate(direction = ifelse(NES>0, "up", "down"))
```

```{r, message=FALSE, warning=FALSE}
DT::datatable(day4gs_stats_all %>% 
              select(c(name, n, NES, padj)), 
              filter = "bottom", 
              caption = "Hallmark pathways affected at 
              Day4 DNL 919 treatment vs ATV:ISO")
```

```{r, fig.width=10,fig.height=4}
ggplot(day4gs_stats, aes(x = reorder(name, NES), y = NES, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_bw() +
  scale_fill_manual(values = c("dodgerblue3", "firebrick3")) +
  theme(legend.position = "none") +
  labs(x = "", y = "GSEA Normalized Enrichment Score")
  
```

## day 7
```{r, message=FALSE, warning=FALSE}
day7gsea <- ffsea(x = Day7vsVeh.top.table, 
                  rank_by = "t", 
                  methods = "fgsea", 
                  fsets = gdb.h)

day7gs_stats_all <- day7gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day7gs_stats <- day7gsea %>% 
    tidy() %>% 
    arrange(desc(NES)) %>% 
    filter(padj < 0.05) %>% 
    transmute(name, n, NES, pval, padj, mean.logFC, mean.t, leadingEdge)

day7gs_stats <- day7gs_stats %>% mutate(direction = ifelse(NES>0, "up", "down"))
```

```{r, message=FALSE, warning=FALSE}
DT::datatable(day7gs_stats_all %>% 
                select(c(name, n, NES, padj)), 
              filter = "bottom", 
              caption = "Hallmark pathways affected at 
              Day7 DNL 919 treatment vs ATV:ISO")
```

```{r, fig.width=10,fig.height=4}
ggplot(day7gs_stats, aes(x = reorder(name, NES), y = NES, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_bw() +
  scale_fill_manual(values = c("dodgerblue3", "firebrick3")) +
  theme(legend.position = "none") +
  labs(x = "", y = "GSEA Normalized Enrichment Score")
  
```

# Metabolic pathways vs ATV:ISO
```{r, message=FALSE, warning=FALSE}
##Get log2(cpm) and sample info. Arrange samples by group.
norm.counts <- log2(cpm(DGEList_filt) + 1)
sample_annotation <- DGEList_filt$samples %>% arrange(group)


##Get the genes from each pathway in the GSEA
oxphos_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_OXIDATIVE_PHOSPHORYLATION"),]$leadingEdge %>% 
  unlist() 
glycolysis_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_GLYCOLYSIS"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
metabolism.counts <- norm.counts[c(oxphos_gsea_genes[2:16], 
                                   glycolysis_gsea_genes[1:15]), 
                                 rownames(sample_annotation)]
gene_names <- DGEList_filt$genes %>% 
  select(c(feature_id, name))

##Change rownames to common gene symbols
rownames(metabolism.counts) <- gene_names[rownames(metabolism.counts),2]


gene_annotation <- data.frame(row.names = rownames(metabolism.counts), 
                              pathway = c(rep("OXPhos", 15), 
                                          rep("Glycolysis", 15)))
column_annotation <- data.frame(row.names = rownames(sample_annotation), 
                                group = sample_annotation$group)

annoCol<-list(pathway=c(OXPhos="Gray20", Glycolysis="Gray60"),
              group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = metabolism.counts,
                   cluster_rows = F,
                   cluster_cols = F,
                   scale= "row",
                   gaps_row = c(15),
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   annotation_row = gene_annotation,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "Metabolism DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```

# Other activation signatures vs ATV:ISO
```{r, message=FALSE, warning=FALSE}
##Get the genes from each pathway in the GSEA
allograft_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_ALLOGRAFT_REJECTION"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
allograft.counts <- norm.counts[allograft_gsea_genes[1:25], 
                                rownames(sample_annotation)]

##Change rownames to common gene symbols
rownames(allograft.counts) <- gene_names[rownames(allograft.counts),2]

annoCol2<-list(group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = allograft.counts,
                   cluster_rows = T,
                   cluster_cols = F,
                   scale= "row",
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   treeheight_row = 0,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "Allograft Rejection DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```


```{r, message=FALSE, warning=FALSE}
##Get the genes from each pathway in the GSEA
G2M_gsea_genes <- day1gs_stats[which(
  day1gs_stats$name == "HALLMARK_G2M_CHECKPOINT"),]$leadingEdge %>% 
  unlist() 

##Make an expression matrix of only the top 15 genes from each pathway
G2M.counts <- norm.counts[G2M_gsea_genes[1:25], rownames(sample_annotation)]

##Change rownames to common gene symbols
rownames(G2M.counts) <- gene_names[rownames(G2M.counts),2]

annoCol2<-list(group = c(Vehicle = "#1B9E77", 
                        ATV_DNP02 = "#D95F02", 
                        DNL919__mgkg10__days_01 = "#7570B3", 
                        DNL919__mgkg10__days_04 = "#E7298A",
                        DNL919__mgkg10__days_07 = "#66A61E"))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
pheatmap::pheatmap(mat = G2M.counts,
                   cluster_rows = T,
                   cluster_cols = F,
                   scale= "row",
                   gaps_col = c(6, 14, 22, 30), 
                   show_colnames = F, 
                   treeheight_row = 0,
                   annotation_col = column_annotation,
                   annotation_colors = annoCol,
                   annotation_names_row = F,
                   annotation_names_col = F, border_color = NA, 
                   main = "G2M Checkpoint DNL919 (hTRME2 KI Mice)", 
                   color = colorRampPalette(colors = c("Blue4", 
                                                       "White", 
                                                       "Red4"))((100)))
```


# Individual Gene plots
```{r, message=FALSE, warning=FALSE}
selected_counts <- norm.counts[c("ENSMUSG00000002602",
                                 "ENSMUSG00000030789", 
                                 "ENSMUSG00000050335", 
                                 "ENSMUSG00000041736"), 
                               rownames(sample_annotation)]
rownames(selected_counts) <- gene_names[rownames(selected_counts),2]
selected_counts <- as.data.frame(t(selected_counts))

selected_counts$group <- plyr::mapvalues(sample_annotation$group, 
                                         from = unique(sample_annotation$group), 
                                         to = c("Vehicle", 
                                                "ATV:ISO", 
                                                "day 1", 
                                                "day 4", 
                                                "day 7"))

selected_counts$treatment <- plyr::mapvalues(sample_annotation$treatment, 
                                         from = unique(
                                           sample_annotation$treatment), 
                                         to = c("Vehicle", 
                                                "ATV:ISO", 
                                                "ATV:TREM2"))

selected_counts$sex <- sample_annotation$sex

```

## Phagocytosis Genes
```{r, message=FALSE, warning=FALSE}
axl <- ggplot(selected_counts, aes(x = group, y = Axl)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Axl")

itgax <- ggplot(selected_counts, aes(x = group, y = Itgax)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Itgax")

lgals3 <- ggplot(selected_counts, aes(x = group, y = Lgals3)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1.2) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Lgals3")


# Faceted by sex
axl_faceted <- ggplot(selected_counts, aes(x = group, y = Axl)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Axl") +
  facet_wrap(~sex)

itgax_faceted <- ggplot(selected_counts, aes(x = group, y = Itgax)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Itgax") +
  facet_wrap(~sex)

lgals3_faceted <- ggplot(selected_counts, aes(x = group, y = Lgals3)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1.2) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Lgals3") +
  facet_wrap(~sex)

```

```{r, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, fig.width=7, fig.height=2.75}
gridExtra::grid.arrange(axl, itgax, lgals3, nrow = 1)
```

Interestingly, Axl levels in response to ATV:ISO were sex-dependent. Namely, 
females have an increase in Axl levels in response to Axl, while males do not 
(except for one outlier). 

```{r, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, fig.width=12, fig.height=2.75}
gridExtra::grid.arrange(axl_faceted, itgax_faceted, lgals3_faceted, nrow = 1)
```

## Tspo
```{r, fig.width=2.34, fig.height=2.75, message=FALSE, warning=FALSE}
ggplot(selected_counts, aes(x = group, y = Tspo)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Tspo") +
  coord_cartesian(ylim=c(5,9))
```

```{r, fig.width=4.5, fig.height=2.75, message=FALSE, warning=FALSE}
ggplot(selected_counts, aes(x = group, y = Tspo)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = "gray30")) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) +
  geom_dotplot(binaxis = "y", aes(fill = treatment), 
               position = position_jitter(width = 0.01, height = 0.1), 
               dotsize = 1) +
  theme_classic() +
  scale_fill_manual(values = c("gray20", "#FD8008", "gray90", "white")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  labs(y = "log2 (cpm + 1)") + 
  ggtitle("Tspo") +
  coord_cartesian(ylim=c(5,9)) +
  facet_wrap(~sex)
```
